---
title: "Code for reproducibility"
subtitle: Paper "Cellwise outlier detection in heterogeneous populations" by Zaccaria, G., García-Escudero, L.A., Greselin, F., Mayo-Íscar, A.- Simulation Study - Main Article - Section 3, Figures 1 and 2
author: "Zaccaria, G."
date: "April 16, 2025"
output: html_document
---

This file reproduces Figures 1 and 2 in Section 3 of the Main Article. Estimated computation time: 5-10 seconds.
To reproduce an entire figure from the simulation study, as reported in the Main Article or in the Supplementary Material, it is necessary to run all scenarios for each level of contamination or at least one scenario of the simulation study for all three levels of contamination (0%, 5%, and 10%). We have attached an .RData file containing the results of the three scenarios reported in the Main Article, called "Data-Figs1-2-Complete.RData", which can be used to fully reproduce Figures 1 and 2.

```{r}
rm(list = ls())
```

Load the necessary packages and the data.

```{r, echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
required_packages <- c("ggplot2", "tidyr", "dplyr", "patchwork")
for (package in required_packages) {
  if (!require(package, character.only = TRUE)) {
    message(paste("Package", package, "is not installed. Attempting to install now."))
    install.packages(package)
  }
  library(package, character.only = TRUE)
}

load("Data-Figs1-2-Complete.RData")
```

Prepare the object for ggplot.

```{r}
df_long <- stat_ggplot %>%
  pivot_longer(cols = c(MR, MSEmu1, MSEmu2, KLsigma1, KLsigma2), 
               names_to = "Metric", 
               values_to = "Value")

df_long_Scen3 <- stat_ggplotscen3 %>%
  pivot_longer(cols = c(MR, MSEmu1, MSEmu2, MSEmu3, MSEmu4, KLsigma1, KLsigma2, KLsigma3, KLsigma4), 
               names_to = "Metric", 
               values_to = "Value")

df_tot <- rbind(df_long, df_long_Scen3)

df_tot$Metric <- factor(df_tot$Metric, levels = c("MR", "MSEmu1", "MSEmu2", "KLsigma1", "KLsigma2", "MSEmu3", "MSEmu4", "KLsigma3", "KLsigma4"))
metric_labels <- c("MR" = "mMR", 
                   "MSEmu1" = "log[10](MSE*mu[1])",
                   "MSEmu2" = "log[10](MSE*mu[2])",
                   "MSEmu3" = "log[10](MSE*mu[3])",
                   "MSEmu4" = "log[10](MSE*mu[4])",
                   "KLsigma1" = "log[10](KL*Sigma[1])",
                   "KLsigma2" = "log[10](KL*Sigma[2])",
                   "KLsigma3" = "log[10](KL*Sigma[3])",
                   "KLsigma4" = "log[10](KL*Sigma[4])")
```

Reproduce Figure 1.

```{r echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
df_tot_MR <- df_tot[which(df_tot$Metric == "MR"), ]
p <- ggplot(df_tot_MR, aes(x = Outlier, y = Value, color = Model, group = Model, shape = Model, linetype = Model)) +
  geom_line() + 
  geom_point() +
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 7, 8, 9)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "1234", "F2")) +
  facet_wrap(vars(Scenario), scales = "free",
          labeller = labeller(Metric = as_labeller(label_parsed))) +  
  scale_x_discrete(breaks = c(0, 5, 10), labels = c(0, 5, 10)) +  
  xlab("% of outlying values") +
  ylab("mMR") +
  theme(strip.text = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "bottom",
        axis.text.x = element_text(angle = 0, hjust = 0.5))
p
```

Reproduce Figure 2.

```{r echo = TRUE, results = 'hide', warning = FALSE, message = FALSE}
df_tot_param <- df_tot[which(df_tot$Metric != "MR"), ]
df_tot_param <- df_tot_param %>%
  mutate(ScenarioMetric = paste(Scenario, Metric, sep = "_"))
df_tot_param$ScenarioMetric <- factor(df_tot_param$ScenarioMetric, levels = c(
  "Scenario 1_MSEmu1", "Scenario 1_MSEmu2", "Scenario 2_MSEmu1", "Scenario 2_MSEmu2",
  "Scenario 3_MSEmu1", "Scenario 3_MSEmu2", "Scenario 3_MSEmu3", "Scenario 3_MSEmu4",
  "Scenario 1_KLsigma1", "Scenario 1_KLsigma2", "Scenario 2_KLsigma1", "Scenario 2_KLsigma2",
  "Scenario 3_KLsigma1", "Scenario 3_KLsigma2", "Scenario 3_KLsigma3", "Scenario 3_KLsigma4"))

df_tot_param <- df_tot_param %>%
    mutate(Type = ifelse(grepl("^MSE", Metric), "MSE", "KL"))
df_MSE <- df_tot_param %>% filter(Type == "MSE")
df_KL  <- df_tot_param %>% filter(Type == "KL")

custom_labels <- c(
  "Scenario 1_MSEmu1" = "Scenario 1 \n g = 1",
  "Scenario 1_MSEmu2" = "Scenario 1 \n g = 2",
  "Scenario 2_MSEmu1" = "Scenario 2 \n g = 1",
  "Scenario 2_MSEmu2" = "Scenario 2 \n g = 2",
  "Scenario 3_MSEmu1" = "Scenario 3 \n g = 1",
  "Scenario 3_MSEmu2" = "Scenario 3 \n g = 2",
  "Scenario 3_MSEmu3" = "Scenario 3 \n g = 3",
  "Scenario 3_MSEmu4" = "Scenario 3 \n g = 4",
  "Scenario 1_KLsigma1" = "Scenario 1 \n g = 1",
  "Scenario 1_KLsigma2" = "Scenario 1 \n g = 2",
  "Scenario 2_KLsigma1" = "Scenario 2 \n g = 1",
  "Scenario 2_KLsigma2" = "Scenario 2 \n g = 2",
  "Scenario 3_KLsigma1" = "Scenario 3 \n g = 1",
  "Scenario 3_KLsigma2" = "Scenario 3 \n g = 2",
  "Scenario 3_KLsigma3" = "Scenario 3 \n g = 3",
  "Scenario 3_KLsigma4" = "Scenario 3 \n g = 4"
)

df_MSE_row1 <- df_MSE %>%
  filter(ScenarioMetric %in% c("Scenario 1_MSEmu1", "Scenario 1_MSEmu2", 
                               "Scenario 2_MSEmu1", "Scenario 2_MSEmu2"))
pMSE_row1 <- ggplot(df_MSE_row1, aes(x = Outlier, y = Value, color = Model, group = Model, shape = Model, linetype = Model)) +
  geom_line() + 
  geom_point() +
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 7, 8, 9)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "1234", "F2")) +
  facet_wrap(vars(ScenarioMetric), scales = "free", nrow = 1, 
             labeller = labeller(ScenarioMetric = as_labeller(custom_labels))) +  
  scale_y_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "", y = expression(MSE*mu[g])) +
  theme(strip.text = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "none")
df_MSE_row2 <- df_MSE %>%
  filter(!ScenarioMetric %in% c("Scenario 1_MSEmu1", "Scenario 1_MSEmu2", 
                                "Scenario 2_MSEmu1", "Scenario 2_MSEmu2"))
pMSE_row2 <- ggplot(df_MSE_row2, aes(x = Outlier, y = Value, color = Model, group = Model, shape = Model, linetype = Model)) +
  geom_line() + 
  geom_point() +
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 7, 8, 9)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "1234", "F2")) +
  facet_wrap(vars(ScenarioMetric), scales = "free", nrow = 1, 
             labeller = labeller(ScenarioMetric = as_labeller(custom_labels))) +  
  scale_y_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "", y = expression(MSE*mu[g]),) +
  theme(strip.text = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "none")

df_KL_row1 <- df_KL %>%
  filter(ScenarioMetric %in% c("Scenario 1_KLsigma1", "Scenario 1_KLsigma2",
                               "Scenario 2_KLsigma1", "Scenario 2_KLsigma2"))
pKL_row1 <- ggplot(df_KL_row1, aes(x = Outlier, y = Value, color = Model, group = Model, shape = Model, linetype = Model)) +
  geom_line() + 
  geom_point() +
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 7, 8, 9)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "1234", "F2")) +
  facet_wrap(vars(ScenarioMetric), scales = "free", nrow = 1, 
             labeller = labeller(ScenarioMetric = as_labeller(custom_labels))) +
  scale_y_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "", y = expression(KL*Sigma[g])) +
  theme(strip.text = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "none") 

df_KL_row2 <- df_KL %>%
  filter(!ScenarioMetric %in% c("Scenario 1_KLsigma1", "Scenario 1_KLsigma2",
                                "Scenario 2_KLsigma1", "Scenario 2_KLsigma2"))
pKL_row2 <- ggplot(df_KL_row2, aes(x = Outlier, y = Value, color = Model, group = Model, shape = Model, linetype = Model)) +
  geom_line() + 
  geom_point() +
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 7, 8, 9)) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash", "1234", "F2")) +
  facet_wrap(vars(ScenarioMetric), scales = "free", nrow = 1, 
             labeller = labeller(ScenarioMetric = as_labeller(custom_labels))) +
  scale_y_continuous(trans = "log10", labels = scales::label_number()) +
  labs(x = "% of outlying values", y = expression(KL*Sigma[g])) +
  theme(strip.text = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "bottom")  

p_combined <- pMSE_row1 / pMSE_row2/ pKL_row1 / pKL_row2
p_combined
ggsave("Figure2.pdf", plot = p_combined, width = 12.5, height = 8.33, units = "in")
print("The figure has been saved as 'Figure2.pdf' to ensure optimal visualization with proper dimensions.")
```
```
